---
title: Getting R Up and Running
date: 05-Oct-2015
output: 
    html_document:
        css: styles.css
---
<style type="text/css">
    td:first-child {
        width: 30%;
    }
</style>

If you're entirely new to R, this document provides a step-by-step manual of how to install R and R Studio on your computer. There is also a basic introduction to the R Studio interface, and a run-through on how to install additional packages (where security safeguards prevent automatic installation).

--------

### 1. Installing R ###

- Go to the [R Website](https://cran.r-project.org/bin/windows/base/)
- Click on the hyperlink to download the latest version of the installer
    + Run the .exe installer and choose the default paths
    + You should not need admin privileges to install R
- Try opening the program
    + Go to Start > All Programs > R
    + The icon should be named "R x64 ..."
    + If you can run the program, exit without saving anything
- You can delete the installer once you've managed to run the program

--------

### 2. Installing R Studio ###

- R Studio is a prettier version of R with a user-friendly
- Go to the [R Studio Website](https://www.rstudio.com/products/rstudio/download/)
- To install R Studio, you will need admin privileges. To get around this:
    + Download the Zip archive (for Windows) and save it into an accessible folder
    + Unzip the folder, open it and then the "bin" folder
    + There will be an application called "rstudio"
    + Create a shortcut to this application on your desktop / startup menu

--------

### 3. Getting a feel of the interface ###
There are 4 parts to the RStudio interface:

1. R scripts (top left)
    + This is where you type in your R "code"
    + You can save your code as .R files which can be opened by most text editors
    + To add a comment (not executed even if run) precede the line with a "#"
    + To run a line of code direct your cursor to the relevant line and press Ctrl + Enter. 
    + Alternatively you may click the "Run" bottom at the top right

2. *Console* (bottom left)
    + The console logs all the "output" from running your code
    + You can enter commands directly in the console but they will not be saved

3. *Workspace* & *History* (top right)
    + __Environment__ tab: Keeps track of all the "objects" you have created during your session
    + __History__ tab: Keeps a log of all the commands that you've run during your session

4. Utilities (bottom right)
    + __Files__ tab: Displays your current <u> working directory</u> and the files within
    + __Plots__ tab: Displays the last graph or chart that you plotted. Use "Export" to save your output as a file.
    + __Packages__ tab: A list of all the packages that you've installed. Packages only need to be installed once to be used. But you need to "load" them for each session by using require("package") or library(package)
    + __Help__ tab: An index of files with information on the various functions

There are numerous online resources for orientation to the interface, such as [this one](http://dss.princeton.edu/training/RStudio101.pdf).

--------

### 4. The Working Directory ###

Where functions in R make use of files other than .R scripts, they will search in your <b><i>Working Directory</i></b> for these files. Think of your directories as rooms in a big house. At any time, there is a person navigating through your house, going from room to room. If you tell the person to access the file Nice_Graph.jpg, the person only searches the room he is currently in for such a file, and will report back with an error if there is none.

To find your current working directory, you can either:

1. Go to the __Files__ tab to see what your working directory currently is, __OR__
2. Run the following command in the console: 

```{r, eval=F}
getwd()
```

To change your current working directory, you can either:

1. Go to the __Files__ tab and navigate to your desired working directory.
    + Click on the 'More' button, then "Set As Working Directory", __OR__
2. Press Ctrl+Shift+H and choose your directory manually, __OR__
3. Run the following command in the console, replacing '~' with your directory name
```{r, eval=F}
setwd("~")
```

--------

### 5. Installing packages ###
Many useful functions in R come in add-ons called 'packages'. R comes with a series of pre-installed basic packages such as 'datasets', 'stats', 'utils', etc. For added functions and tools, you can install other packages made available on the CRAN network.

If you run R on your own personal computer, the easiest way to install packages is to type a single line of code into the console. This accesses an online 'repository' containing the files and downloads them to your computer. For instance to install the package __shiny__ you would type:

```{r, eval=F}
install.packages("shiny")
```

However, this method will not work on your work laptop. Instead, you will need to perform a manual installation. The next series of steps shows you how to do this:

1. Access the package website on the CRAN network. For example, for the package 'htmltools' click [here](https://cran.r-project.org/web/packages/htmltools/index.html)
    + You can replace __htmltools__ in the address with the package of your choice

2. Check if your package has any 'dependencies'
    + __Imports__: lists the prerequisites required for successful installation of your desired package
    + For the package [RColorBrewer]((https://cran.r-project.org/web/packages/RColorBrewer/index.html)) there are no imports, so we can go ahead and install it
    + For the package [htmltools](https://cran.r-project.org/web/packages/htmltools/index.html) there are 2 dependencies: __utils__ (pre-installed) and __digest__
    + If there are dependencies, you have to install the 'dependent' packages first

3. Under __Downloads__, click on one of the __Windows binaries__ to download it

4. Set your working directory to the Downloads folder on your computer

5. Finally, type the following code into your console, replacing 'htmltools_0.2.6.zip' with the <u> name of the downloaded, unzipped folder</u>:

```{r, eval=F}
install.packages(pkgs="htmltools_0.2.6.zip", repos = NULL, type="source")
```

### 6. Short Familiarisation Exercise ###

- Create a new R script and save it in your Documents folder as "Test.R"
- Type in each of the following lines of code into the R script, then run them one by one
- Observe what happens to the *console* and *environment*

Code       | Notes
-----------|-------------
`x <- 2`   | Assign the value of 2 to an object called `x`
`3 + x`    | Add 3 to the object `x`
`y <- c(1, 3, 5, 7, 9)`| Create a *vector* of values called `y`
`y[3]`     | Print the 3rd element in the *vector* `y`
`y[3:5]`   | Print the 3rd to 5th elements in the *vector* `y`
`y*2`      | Multiply the *vector* `y` by 2, then print the output
`y*x`      | Multiply the *vector* `y` by the *scalar* `x`, then print the output
`X <- y*x` | Save the result from the previous line of code into an object `X`
`X == x`   | Check if the object `X` is equal to the object `x`
`animals <- data.frame(`<br>&nbsp;&nbsp;&nbsp;&nbsp;`type=c("Cat", "Dog", "Elephant"),`<br>&nbsp;&nbsp;&nbsp;&nbsp;`name=c("Cathy", "Donald", "Ervin"))` | Create a *data frame* with 2 columns, <br> *type* and *name*, and save it <br> as an object called `animals`
`names(animals)`| Display the column names of the *data frame* called `animals`
`animals[,1]` | Display the first column of `animals`
`animals[,1]` | Display the first row of `animals`

